Trump vs. Harris: A Data-Driven Forecast modeling the 2024 US Presidential Elections

### 2024-10-28 paper/paper.qmd
Update Visualizations and clean up packages enviroment set up

---
title: "Trump vs. Harris: A Data-Driven Forecast modeling the 2024 US Presidential Elections"
author: 
  - Sophia Brothers
  - Deyi Kong
thanks: "Code and data are available at: [https://github.com/eeeee-cmd/US_Election/](https://github.com/eeeee-cmd/US_Election/)."
date: today
date-format: long
abstract: "This study models the 2024 U.S. Presidential election outcomes using advanced statistical techniques and predictive modeling. The analysis incorporates data from various polling sources and considers key demographic and political factors to forecast both the popular vote and Electoral College results. The findings suggest a close race between the two main candidates, Donald Trump and Kamala Harris, with a predicted victory for Trump in the Electoral College. This analysis contributes to understanding how public opinion data and electoral systems interact, providing insights into electoral dynamics in contemporary U.S. politics."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The data used in this analysis comes from a combination of publicly available polling data for the 2024 U.S. Presidential election. The analysis leverages R and several libraries, including tidyverse [@tidyverse], janitor [@janitor], knitr [@knitr], dplyr [@dplyr], purrr [@purrr], and here [@here], for data manipulation, and ggplot2 [@ggplot] for visualization. The dataset covers various polls conducted across multiple states, capturing the support for each major candidate—Donald Trump and Kamala Harris—along with detailed attributes of the polls.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(here)
library(arrow)

@@

print(results)
```

@@

library(tidyverse)
library(here)
library(arrow)

@@

print(electoral_totals)

state_winners <- predictions_with_votes %>%
  select(State, TrumpPredicted, HarrisPredicted) %>%
  mutate(
    Winner = case_when(
      TrumpPredicted > HarrisPredicted ~ "Donald Trump",
      HarrisPredicted > TrumpPredicted ~ "Kamala Harris",
      TRUE ~ "Tie"
    )
  )

print(state_winners, n=54)
```

@@

```{r}
#| label: fig-sum
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
cleaned_data <- read_parquet(here::here("data/analysis_data/president_polls_cleaned.parquet"))

@@

print(summary_stats)
```

```{r}
#| label: fig-pollscore
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = PollScore)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Poll Scores", x = "Poll Score", y = "Frequency")
```

```{r}
#| label: fig-samplesize
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = SampleSize)) +
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  labs(title = "Distribution of Sample Size", x = "Sample Size", y = "Frequency")
```

```{r}
#| label: fig-percentage
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = Percentage)) +
  geom_histogram(bins = 30, fill = "orange", alpha = 0.7) +
  labs(title = "Distribution of Percentage", x = "Percentage", y = "Frequency")
```

```{r}
#| label: fig-candidatepercentage
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = CandidateName, y = Percentage)) +
  geom_boxplot() +
  labs(title = "Percentage by Candidate", x = "Candidate", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

will update these box plots to different visuals due to box plots not showing the full dataset!

```{r}
#| label: fig-statepercentage
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = State, y = Percentage)) +
  geom_boxplot() +
  labs(title = "Percentage by State", x = "State", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig-corrmatrix
#| fig-cap: [Need Edit]
#| fig-subcap: [""]
#| echo: false
#| eval: true
#| warning: false
#| message: false
numerical_data <- cleaned_data %>%
  select(PollScore, SampleSize, Percentage)

>>>>

---
title: "Trump vs. Harris: A Data-Driven Forecast modeling the 2024 US Presidential Elections"
subtitle: "A close race but Trump, trumps Harris"
author: 
  - Sophia Brothers
  - Deyi Kong
  - Rayan Awad Alim
thanks: "Code and data are available at: [https://github.com/eeeee-cmd/US_Election/](https://github.com/eeeee-cmd/US_Election/)."
date: today
date-format: long
abstract: "This study models the 2024 U.S. Presidential election outcomes using advanced statistical techniques and predictive modeling. The analysis incorporates data from various polling sources and considers key demographic and political factors to forecast both the popular vote and Electoral College results. The findings suggest a close race between the two main candidates, Donald Trump and Kamala Harris, with a predicted victory for Trump in the Electoral College. This analysis contributes to understanding how public opinion data and electoral systems interact, providing insights into electoral dynamics in contemporary U.S. politics."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# Install Packages if not downloaded:
required_packages <- c(
  "tidyverse", "sf", "here", "arrow", "ggplot2", "janitor", "purrr", "knitr", "kableExtra"
) 
for (p in required_packages) { 
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(tidyverse)
library(sf)
library(dplyr)
library(here)
library(arrow)
library(ggplot2)
library(janitor)
library(purrr)
library(knitr)
library(kableExtra)
```

@@

The data used in this analysis comes from a combination of publicly available polling data for the 2024 U.S. Presidential election[@FiveThirtyEightPolls]. The analysis leverages the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `purrr` [@purrr], and `here` [@here], for data manipulation, and `ggplot2` [@ggplot] for visualization. The dataset covers various polls conducted across multiple states, capturing the support for each major candidate—Donald Trump and Kamala Harris—along with detailed attributes of the polls.

@@

@@

results %>%
  kable(
    caption = "Predicted Percentage Vote for Donald Trump and Kamala Harris",
    col.names = c("Candidate Name", "Average Predicted Percentage", "Normalized Percentage"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

@@

electoral_totals %>%
  kable(
    caption = "Predicted Number of Electoral Votes for Donald Trump and Kamala Harris",
    col.names = c("Trumps Total Electoral Votes", "Harriss Total Electoral Votes"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

state_winners <- predictions_with_votes %>%
  select(State, TrumpPredicted, HarrisPredicted) %>%
  mutate(
    Winner = case_when(
      TrumpPredicted > HarrisPredicted ~ "Donald Trump",
      HarrisPredicted > TrumpPredicted ~ "Kamala Harris",
      TRUE ~ "Tie"
    )
  )
#print(state_winners, n=54)

state_winners %>%
  slice(1:53) %>%
  kable(
    caption = "Predicted Percentage of Electoral Votes Per State",
    col.names = c("State", "Trump Predicted %", "Harris Predicted %", "Winner"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| label: tbl-sum
#| fig-cap: [Need Edit]
#| echo: false
#| eval: true
#| warning: false
#| message: false
cleaned_data <- read_parquet(here::here("data/analysis_data/president_polls_cleaned.parquet"))

@@

# print(summary_stats)

summary_stats %>%
  kable(
    caption = "Summary Statistics of Model Results",
    col.names = c("Average Poll Score", "Average Sample Size", "Average Percentage", "Total Polls"),
    digits = 3,
    booktabs= TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| label: fig-predictedscoremap
#| fig-cap: "Mapped Model-predicted outsom for 2024 elections by State"
#| echo: false
#| eval: true
#| warning: false
#| message: false

states_map <- map_data("state")

electoral_data <- predictions_with_votes
electoral_data$State <- tolower(electoral_data$State) # matching state names

merged_data <- inner_join(states_map, electoral_data, by = c("region" = "State"))

map <- ggplot(merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Winner), color = "white") +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue")) +
  labs(title = "2024 US Presidential Election Predictions by State",
       fill = "Predicted Winner") +
  theme_void() +
  theme(legend.position = "right")

map
```

```{r}
#| label: fig-pollscoredistribution
#| fig-cap: Distribution of Poll Scores by Candidate with denisty overlayed
#| echo: false
#| eval: true
#| warning: false
#| message: false

Select_candidates_cleaned_data <- cleaned_data %>%
  filter(CandidateName %in% c("Donald Trump", "Kamala Harris"))

ggplot(Select_candidates_cleaned_data, aes(x = PollScore)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "blue",
                 color = "black",
                 alpha = 0.7) +
    geom_density(color = "black", fill = "salmon", alpha = 0.6) +
  labs(title = "Distribution of Poll Scores by Candidate",
       x = "Poll Score", 
       y = "Frequency") +
  theme_minimal() +
  facet_wrap(~ CandidateName)
```

```{r}
#| label: fig-samplesize
#| fig-cap: [Need Edit]
#| echo: false
#| eval: true
#| warning: false
#| message: false

ggplot(cleaned_data, aes(x = SampleSize)) +
  geom_histogram(bins = 30,
                 fill = "green",
                 color = "black",
                 alpha = 0.7) +
  labs(title = "Distribution of Sample Size",
       x = "Sample Size", 
       y = "Frequency")+
  theme_minimal()
```

```{r}
#| label: fig-percentage
#| fig-cap: [Need Edit]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = Percentage)) +
  geom_histogram(bins = 30,
                 fill = "orange",
                 color = "black",
                 alpha = 0.7) +
  labs(title = "Distribution of Percentage", 
       x = "Percentage",
       y = "Frequency") +
  theme_minimal()
```

```{r}
#| label: fig-candidatepercentage
#| fig-cap: [Need Edit]
#| echo: false
#| eval: true
#| warning: false
#| message: false

harris_trump_selection <- cleaned_data %>%
  filter(CandidateName %in% c("Donald Trump", 
                              "Kamala Harris"))

all_candidates <- cleaned_data %>%
  mutate(CandidateName = "All Candidates")

combined_data <- bind_rows(
  harris_trump_selection,
  all_candidates
)

ggplot(combined_data, aes(x = CandidateName, 
                          y = Percentage, 
                          fill = CandidateName)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.1, color = "black", outlier.shape = NA) +
  labs(title = "Percentage by Candidate", 
       x = "Candidate", 
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("grey", "salmon", "blue")) +
  facet_wrap(~CandidateName, scales = "free_x") +
  theme_minimal()
```

will update these box plots to different visuals due to box plots not showing the full dataset! (#Note to team: updated to violin plots to show distribution and selected only trump and harris and another one for all the candidates to show distribution of dataset)

```{r}
#| label: fig-statepercentage
#| fig-cap: [Need Edit]
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(cleaned_data, aes(x = State, y = Percentage)) +
  geom_boxplot() +
  labs(title = "Percentage by State", x = "State", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig-corrmatrix
#| fig-cap: [Need Edit]
#| echo: false
#| eval: true
#| warning: false
#| message: false
numerical_data <- cleaned_data %>%
  select(PollScore, SampleSize, Percentage)
